A Linear Solution of the Linearity Challenge

### 2025-12-28 paper/related.tex
revised related

    \caption{\label{tab:comparison} A summary of different formalisations of

@@

\Cref{tab:comparison} reports a summary of the aforementioned formalisations
(sorted by publication date) and also our own. The first five columns identify
the calculus being formalised specifying its reference paper, the used proof
assistant and an acronym that gives an idea of the overall flavour of the
calculus.
%
The acronyms \CP~\cite{Wadler14,GayVasconcelos25} and
\GV~\cite{GayVasconcelos10,Wadler14} are well-known in the literature on session
types. The acronym \SessionPI refers to (variants of) the session-based
$\pi$-calculus presented by \citet{HondaVasconcelosKubo98} while \LinearPI
refers to (variants of) the linear
$\pi$-calculus~\cite{KobayashiPierceTurner99}. Finally, \SCP is the structural
version of \CP introduced by \citet{SanoKavanaghPientka23} and \Calculus is our
calculus.
%
We emphasize that the actual calculus being formalised usually differs from
(typically, is a strict subset of) the one identified by the acronym and that
the same acronym may refer to different versions of the same calculus.
%
We also indicate whether the calculus features \emph{cuts} (that is, the
combination of restriction and parallel composition in correspondence with the
cut of linear logic) and continuations.

Concerning the use of continuations, the approaches based directly on the linear
$\pi$-calculus (into which sessions can be encoded) are marked with {\YES} and
those based on a calculus with native sessions are marked with {\NO}. The
calculus \SCP is marked with {\MAYBE} because, while not directly inspired to
the linear $\pi$-calculus, it makes use of explicit continuations for defining
the predicates that check the linear usage of channels. Finally, all the
approaches based on \GV are also marked with \MAYBE. In principle, \emph{none}
of these calculi makes use of continuations. However, \GV is designed in such a
way that each operation acting on a channel $s$ is a function that returns the
result of the operation (if any) \emph{along with the same channel $s$}. In this
way, the type of $s$ can be conveniently ``updated'' to take into account the
effect of the operation. As observed by \citet{Padovani17}, this semantics of
the communication primitives is virtually indistinguishable from one making use
of explicit continuation channels.

>>>>

    \caption{\label{tab:comparison} Overview of different formalisations of

@@

\Cref{tab:comparison} shows an overview of the aforementioned formalisations
(sorted by publication date) including our own. The first five columns identify
the calculus being formalised. We provide its reference paper, the language in
which it is formalised and an acronym that gives an idea of the flavour of the
calculus. We also specify whether the calculus features \emph{cuts} (that is,
the combination of restriction and parallel composition that corresponds with
the cut of linear logic) and continuations.
%
\CP~\cite{Wadler14,GayVasconcelos25} and \GV~\cite{GayVasconcelos10,Wadler14}
are well-known acronyms in the literature on session types. \SessionPI refers to
(variants of) the session-based $\pi$-calculus presented by
\citet{HondaVasconcelosKubo98} while \LinearPI refers to (variants of) the
linear $\pi$-calculus~\cite{KobayashiPierceTurner99}. Finally, \SCP is the
structural version of \CP introduced by \citet{SanoKavanaghPientka23} and
\Calculus is our calculus.
%
We emphasize that the actual calculus being formalised usually differs from
(typically, is a strict subset of) the one identified by the acronym and that
the same acronym may refer to different versions of the same calculus. In
particular, \GV in a logical setting is described by \citet{Wadler14} but its
first (non-logical) version is due to \citet{GayVasconcelos10}.

Concerning the use of continuations, the approaches based directly on the linear
$\pi$-calculus (into which sessions can be encoded) are marked with {\YES} and
those based on a calculus with native sessions are marked with {\NO}. The
calculus \SCP is marked with {\MAYBE} because, while not directly inspired to
the linear $\pi$-calculus, it makes use of explicit continuations for defining
the predicates that check the linear usage of channels. Finally, all the
approaches based on \GV are also marked with \MAYBE. In principle, \emph{none}
of these calculi makes use of continuations. However, \GV is designed in such a
way that each operation acting on a channel $s$ is a function that returns the
result of the operation (if present) \emph{along with the same channel $s$}. In
this way, the type of $s$ can be conveniently ``updated'' to take into account
the effect of the operation. As observed by \citet{Padovani17}, this semantics
of the communication primitives is virtually indistinguishable from one making
use of explicit continuation channels.
